---
title: "Effects of harmful weather events on population health and economic consequences"
author: "Isabelle You"
date: "September 7, 2017"
output: pdf_document
---



The basic goal of this project is to explore the NOAA Storm Database and the following questions about severe weather events:

1. Across the United Sates, which types of events are the most harmful with respect to population health?
2. Across the united States, which types of events have the greatest economic consequences?

## Load and clean-up data

data source: https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2

### Download the NOAA data from website and read into R

```{r}
# load data into R

rm(list=ls())

library(ROracle)
library(DBI)


if (!file.exists("coursera")){dir.create("coursera")}
fileurl="https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
download.file(fileurl,destfile="~/R/coursera/stormdata.csv.bz2")
datedownloaded=date()

storm=read.csv("~/R/coursera/stormdata.csv.bz2", header=T, sep=",")
```

### Clean-up data

```{r}
# clea-up the data

library(dplyr)

# change storm type all in upper letters

storm$EVTYPE=toupper(storm$EVTYPE)





storm=storm %>% 
        mutate (DEATH_INJURIES=FATALITIES+INJURIES) %>%
        mutate(PROPDMG1=ifelse(PROPDMGEXP %in% c("K", "k"), PROPDMG*1000,
                               ifelse(PROPDMGEXP %in% c("m", "M"), PROPDMG*1000000,
                                      ifelse(PROPDMGEXP %in% c("B","b"), PROPDMG*1000000000, PROPDMG)))) %>%
        mutate(PROPDMG_K=PROPDMG1/1000) %>%
        mutate(CROPDMG1=ifelse(CROPDMGEXP %in% c("K", "k"), CROPDMG*1000,
                               ifelse(CROPDMGEXP %in% c("m", "M"), CROPDMG*1000000,
                                      ifelse(CROPDMGEXP %in% c("B","b"), CROPDMG*1000000000, CROPDMG)))) %>%
        mutate(CROPDMG_K=CROPDMG1/1000) %>%
        mutate(TOTALDMG_K=PROPDMG_K+CROPDMG_K)
```

## Data analysis

### Which types of events are most harmful with respect to pupulation health

```{r}
# Top 10 most harmful events for fatalities

T_death=storm %>%
        group_by(EVTYPE) %>%
        summarise(Total_death=sum(FATALITIES, na.rm=T)) %>%
        arrange(desc(Total_death))

Top10_death=head(T_death, n=10)

print(Top10_death)
```

```{r}
# Top 10 most harmful events for injuries

T_injury=storm %>%
        group_by(EVTYPE) %>%
        summarise(Total_injury=sum(INJURIES, na.rm=T)) %>%
        arrange(desc(Total_injury))

Top10_injury=head(T_injury, n=10)

print(Top10_injury)
```

```{r}
# Top 10 most harmful events for death and injuries

T_death_injury=storm %>%
        group_by(EVTYPE) %>%
        summarise(Total_death_injury=sum(DEATH_INJURIES, na.rm=T)) %>%
        arrange(desc(Total_death_injury))

Top10_death_injury=head(T_death_injury, n=10)

print(Top10_death_injury)
```


### create a bar plot to show the top 20 most harmful weather events for fatalities

```{r}
library(ggplot2)

Top20_death=head(T_death, n=20)

              
      g=ggplot(Top20_death, aes(x=EVTYPE, y=Total_death))
      g+ geom_bar(position="dodge", stat="identity", col="blue") +
              coord_flip() +
              xlab("Weather Event") +
              ylab("Total Number of Fatalities ") +
              ggtitle("The Top 20 Most Harmful Weather Events for Fatalites ")
```
      
      

### Which types of events have the greatest economic consequeces 
 
```{r}
# Top 10 most harmful  events for property damage in thousand dollars
      
      T_propdmg=storm %>%
              group_by(EVTYPE) %>%
              summarise(Total_propdmg=sum(PROPDMG_K, na.rm=T)) %>%
              arrange(desc(Total_propdmg))
      
      Top10_propdmg=head(T_propdmg, n=10)
      
      print(Top10_propdmg)
```

```{r}
# Top 10 most harmful  events for crop damage in thousand dollars
      
      T_cropdmg=storm %>%
              group_by(EVTYPE) %>%
              summarise(Total_cropdmg=sum(CROPDMG_K, na.rm=T)) %>%
              arrange(desc(Total_cropdmg))
      
      Top10_cropdmg=head(T_cropdmg, n=10)
      
      print(Top10_cropdmg)
```

```{r}
# Top 10 most harmful  events for property and crop damage in thousand dollars
      
      T_totaldmg=storm %>%
              group_by(EVTYPE) %>%
              summarise(Totaldmg=sum(TOTALDMG_K, na.rm=T)) %>%
              arrange(desc(Totaldmg))
      
      Top10_totaldmg=head(T_totaldmg, n=10)
      
      print(Top10_totaldmg)
```    

### create a bar plot to show the top 20 most harmful weather events for property and crop damage

```{r}
T_totaldmg=T_totaldmg %>%
        mutate(Totaldmg_in_M=Totaldmg/1000)

      
Top20_totaldmg=head(T_totaldmg, n=20)      
      
      g=ggplot(Top20_totaldmg, aes(x=EVTYPE, y=Totaldmg_in_M))
      g+ geom_bar(position="dodge", stat="identity", col="green") +
              coord_flip() +
              xlab("Weather Event") +
              ylab("Total Damage in Million Dollars ") +
              ggtitle("The Top 20 Most Harmful Weather Events for Property and Crop Damage")
```

## Conclusion

The severe weather events caused lots of deaths and injuries and hugh property and crop damage. The top 3 harmful weather events for fatalities were tornado, excessive heat, and flash flood. The 3 worst weather events caused injuries were tornado, TSTM wind, and flood. Compared to other severe weather events, tornada caused far more deaths and injuries in the United States. Top 3 severe weather events  for property damage were Flood, hurricane/typhoon, and flood. Top 3 harmful weather events for crop damage were drought, flood and river flood. Flood, hurricane/typhoon, and tornada were the 3 morst events for property and crop damage.
